---
title: "Prediction Model In Class"
author: "Md Boshirul Azad"
date: "Fall 2025"
format:
  html: 
    theme: cosmo
    fontsize: .85em
    linestretch: 1.5
    toc: true
    toc-expand: 2
    toc-location: body
    toc-title: Table of Contents
execute:
  echo: false
  warning: false
---

```{r Themes}
# To change look and feel, you can simply adjust themes. 
# Here is a reference. Simply replace 'simples' with one of these names:
# https://quarto.org/docs/output-formats/html-themes.html

```

# I. Executive Summary

## Define Focus of Prediction \

\ What are you trying to predict?\
Ans: We are trying to predict house sales price.

\ What type of prediction model is it?\
Ans: It is linear regression prediction model.

## Communicate Findings \

\ What were your findings from your prediction model?\

Findings:

- Overall the model performs well.

- Predictions are quite close to actual price.

- Accuracy drops when it comes to higher price range.

- Most points follow the diagonal line.

- Neighborhood, overall quality, and Ground living area are the most influential predictors.

\ How do these findings tie back to question or problem at hand?\

Findings:

- This model effectively predicts house prices.

- The model shows a comprehensible data trend even though the accuracy drops at higher price range.

## Take Action 

\Based on the analysis, what do actions do you recommend\

Actions:(Good models)

- Add more variables with higher prices to get more accurate result on higher price range.

- Increase the sample size to increase the overall accuracy of the model.

- Try models with different predictors.

Actions:(Bad model, lm_t5)

- Use more relevant predictors to make model.

- Avoid too simple variables.


# II. Prediction Model

\ Summarization of the model\

- The outcome variable is sale price. 

-The models consists of Neighborhood, Overall quality, Ground living area, Garage cars, lot area. 

- The good models includes Neighborhood, Overall quality, and Ground living area.

- The bad model has Garage cars and lot area.

Overall, the models are accurate and can predict house prices with the exception of high price range houses.

## Collect Data \

\ We are using the Ames Housing data.

### Clear Environment

```{r Clear Environment}
#Clear the environment (be careful with this)
rm(list = ls())

```

### Load Libraries and Custom Functions

```{r Load Libraries}


# Load Libraries
library(ggplot2) # This library is useful for data visualization
library(scales) #Formats output data
library(janitor) #Formats output data
library(tidyverse)
library(forcats)
library(rlang)
library(patchwork)
library(gt)
library(glmnet)
library(tidymodels)  # modern unified framework (includes rsample, parsnip, yardstick, etc.)
library(broom)       # tidy model summaries
library(performance) # model diagnostics
library(see)         # visualization for performance package
library(vip)         # variable importance plots
library(MASS)
library(car)        # for VIF
library(caret)      # for preprocessing and modeling
library(dplyr) # This library is useful for data manipulation
library(conflicted)

conflict_prefer("select", "dplyr")
# tidymodels: the best unified ecosystem for modern R modeling.
# rsample: train/test splitting.
# yardstick: performance metrics.
# performance + see: model diagnostics and visualization.
# vip: interpretable variable importance plots.
# broom: tidy coefficient summaries.

#source("CS217_Functions.R")

```

### Load Data

```{r Load Data}
# Load data set 

# KEY: Make sure the file is in your default workdirectory of your project
#sales_data <- read.csv("sales_transactions.csv", header = T) 
library(AmesHousing)
ames_data <- make_ames()
# glimpse(ames)

```

## Clean Data \

### Convert Data to Useful Definitions

```{r}

## Clean and organize the data better



```

### Modify Data - Create Calculated Fields

We will build two different models with the Ames housing data set. The first model will use regression to predict the house 
price - a continuous value. The 2nd


```{r}
# 1) GET THE SIGNIFICANT FACTOROS

set.seed(123)
split_reg <- initial_split(ames_data, prop = 0.7)
train_data <- training(split_reg)
test_data  <- testing(split_reg)


```
## MAKE THE MODEL TO GET THE SIGN FACTORS
```{r}
predictor_cols <- ames_data %>%
  select(-Sale_Price) %>% #Dont inlclude Sale Price, it is the outcome variable
  select(where(is.numeric)) %>% #Select all numeric predictor variables
  names() #pull just the column names
```
## CHECK THE WHOLE DATA
```{r}
lm_all <- lm(Sale_Price~ ., data = train_data)

# summary(lm_all)
```


## Model 1: Linear Regression Data


```{r Calcuate Fields}

# Linear regression target
ames_reg <- ames_data %>%
  dplyr::select(Sale_Price,
                Gr_Liv_Area, 
                Year_Built, 
                Garage_Cars, 
                Full_Bath, 
                Neighborhood,
                Lot_Area,
                MS_SubClass,
                Street,
                Land_Contour,
                Land_Slope,
                Condition_1,
                Condition_2,
                Overall_Qual,
                Overall_Cond,
                Exterior_1st,
                Mas_Vnr_Area,
                Mas_Vnr_Type,
                Foundation,
                Bsmt_Qual,
                Bsmt_Exposure,
                BsmtFin_Type_2,
                Bsmt_Unf_SF)


```

Handle Missing Values

```{r}
ames_reg$age <- year(today()) - ames_reg$Year_Built

ames_reg$Year_Built <- NULL 

ames_reg$age
```


```{r}
predictor_cols <- ames_reg %>%
  select(-Sale_Price) %>% #Dont inlclude Sale Price, it is the outcome variable
  select(where(is.numeric)) %>% #Select all numeric predictor variables
  names() #pull just the column names
```

### Normalize Predictor Variables 

This is done so that large numeric values (like lot area) doesn't overshadow the impact of small number (like Full Bath). This puts all numerical values on a standard scale. 

```{r}


predictor_cols <- ames_reg %>%
  select(-Sale_Price) %>% #Dont inlclude Sale Price, it is the outcome variable
  select(where(is.numeric)) %>% #Select all numeric predictor variables
  names() #pull just the column names

# Put in a new data set that can be matched back to the original data set for interpretation

ames_reg_clean <- ames_reg %>%
  mutate(across(all_of(predictor_cols), ~ as.numeric(scale(.x))))

```


### Create the Training & Test Data Sets

Training data sets are to develop a base model. This is typically 60-80% of the data set. The reason is to have large enough data set to build a quality model but without "overfitting". 

Test data set is used to prove out the quality of the model. This is typically 40-20% of the original data set.

Overfitting is when you over engineer your model to the data set only and not for general purpose usage across data sets.

```{r}

# 1) TRAIN/SPLIT THE DATA

set.seed(123)
split_reg <- initial_split(ames_reg_clean, prop = 0.7)
train_data <- training(split_reg)
test_data  <- testing(split_reg)


```
###Check for the Significant factors

```{r}
lm_all_clean <- lm(Sale_Price~ ., data = train_data)

# summary(lm_all_clean)
```


### Test some Models to find the significant ones

```{r}

# Define a linear regression model

# Dependent - Y -  Sales_Price - Outcome variable that you are trying to predict
# Independent Variables - f(x) - Input variables, coefficients, predictors, etc. 



lm_2 <- lm(Sale_Price ~ Full_Bath, data = train_data) # Low R score

# summary(lm_2)

lm_3 <- lm(Sale_Price ~ Gr_Liv_Area, data = train_data) # Low R score

# summary(lm_3)

lm_4 <- lm(Sale_Price ~ Garage_Cars, data = train_data) # Low R score

# summary(lm_4)


lm_5 <- lm(Sale_Price ~ Lot_Area, data = train_data) # Low R score

# summary(lm_5)

lm_6 <- lm(Sale_Price ~ Neighborhood, data = train_data) # R = 0.5677

# summary(lm_6)

lm_7 <- lm(Sale_Price ~ age, data = train_data) # Low R score

# summary(lm_7)

lm_8 <- lm(Sale_Price ~ Street, data = train_data) # Low R score

# summary(lm_8)

lm_9 <- lm(Sale_Price ~ Land_Contour, data = train_data) # Low R score

# summary(lm_9)

lm_10 <- lm(Sale_Price ~ Land_Slope, data = train_data) # Low R score

# summary(lm_10)

lm_11 <- lm(Sale_Price ~ Condition_1, data = train_data) # Low R score

# summary(lm_11)

lm_12 <- lm(Sale_Price ~ Condition_2, data = train_data) # Low R score

# summary(lm_12)

lm_13 <- lm(Sale_Price ~ Overall_Qual, data = train_data) # R = 0.708

# summary(lm_13)

lm_14 <- lm(Sale_Price ~ Overall_Cond, data = train_data) # Low R score

# summary(lm_14)

# lm_15 <- lm(Sale_Price ~ Year_Built, data = train_data) # Low R score
# 
# summary(lm_15)

# lm_16 <- lm(Sale_Price ~ Roof_Matl, data = train_data) # Low R score
# 
# summary(lm_16)

lm_16 <- lm(Sale_Price ~ Exterior_1st, data = train_data) # Low R score

# summary(lm_16)

lm_17 <- lm(Sale_Price ~ Mas_Vnr_Type, data = train_data) # Low R score

# summary(lm_17)

lm_18 <- lm(Sale_Price ~ Mas_Vnr_Area, data = train_data) # Low R score

# summary(lm_18)

lm_18 <- lm(Sale_Price ~ Foundation, data = train_data) # Low R score

# summary(lm_18)

lm_19 <- lm(Sale_Price ~ Bsmt_Qual, data = train_data) # R = 0.5283

# summary(lm_19)

lm_20 <- lm(Sale_Price ~ Bsmt_Exposure, data = train_data) # Low R score

# summary(lm_20)

lm_21 <- lm(Sale_Price ~ BsmtFin_Type_2, data = train_data) # Low R score

# summary(lm_21)

lm_22 <- lm(Sale_Price ~ Bsmt_Unf_SF, data = train_data) # Low R score

# summary(lm_22)


lm_togather <- lm(Sale_Price ~ Neighborhood + Overall_Qual + Bsmt_Qual, data = train_data)

# summary(lm_togather)

lm_t1 <- lm(Sale_Price ~ Neighborhood + Overall_Qual, data = train_data)

# summary(lm_t1)

lm_t2 <- lm(Sale_Price ~ Neighborhood + Bsmt_Qual, data = train_data)

# summary(lm_t2)

lm_t3 <- lm(Sale_Price ~ Overall_Qual + Bsmt_Qual, data = train_data)

# summary(lm_t3)

lm_t4 <- lm(Sale_Price ~ Overall_Qual + Gr_Liv_Area, data = train_data)

# summary(lm_t4)


lm_t5 <- lm(Sale_Price ~ Lot_Area + Garage_Cars, data = train_data)

# summary(lm_t5)


```

### Build and Evaluate 3 Distinct Models

```{r}

lm_t1 <- lm(Sale_Price ~ Neighborhood + Overall_Qual, data = train_data)

summary(lm_t1)

lm_t4 <- lm(Sale_Price ~ Overall_Qual + Gr_Liv_Area, data = train_data)

summary(lm_t4)


lm_t5 <- lm(Sale_Price ~ Lot_Area + Garage_Cars, data = train_data)

summary(lm_t5)
```


### Make a prediction and evaluate model quality

The prediction uses the fitted model (meaning a model derived from the training data set) and makes a prediction from the Test data set.

```{r}

# Predict on test data (Neighborhood and overall quality)

predictions <- predict(lm_t1, test_data) 

# Add predicted values to the test data set
test_data$Predicted_Sale_Price <- predictions

# Show the difference in values
test_data$Predicted_vs_Actual <- round(test_data$Sale_Price - test_data$Predicted_Sale_Price,2)

# Evaluate metrics

metrics <- metrics(test_data, truth = Sale_Price, estimate = Predicted_Sale_Price)

metrics$.estimate <- round(metrics$.estimate,2)

test_data %>%
ggplot(aes(x = Sale_Price, y = Predicted_Sale_Price)) +
  geom_point(alpha = 0.6) + #Size of the points
  geom_abline(slope = 1, intercept = 0, color = "red") + #line to show slope
  labs(
    title = "Predicted vs Actual Sale Price",
    x = "Actual Sale Price",
    y = "Predicted Sale Price"
  ) +
  theme_minimal()



```

```{r}
# Predict on test data (Overall_Qual + Gr_Liv_Area)

predictions <- predict(lm_t4, test_data) 

# Add predicted values to the test data set
test_data$Predicted_Sale_Price <- predictions

# Show the difference in values
test_data$Predicted_vs_Actual <- round(test_data$Sale_Price - test_data$Predicted_Sale_Price,2)

# Evaluate metrics

metrics <- metrics(test_data, truth = Sale_Price, estimate = Predicted_Sale_Price)

metrics$.estimate <- round(metrics$.estimate,2)

test_data %>%
ggplot(aes(x = Sale_Price, y = Predicted_Sale_Price)) +
  geom_point(alpha = 0.6) + #Size of the points
  geom_abline(slope = 1, intercept = 0, color = "red") + #line to show slope
  labs(
    title = "Predicted vs Actual Sale Price",
    x = "Actual Sale Price",
    y = "Predicted Sale Price"
  ) +
  theme_minimal()

```

```{r}
# Predict on test data (Lot_Area + Garage_Cars)

predictions <- predict(lm_t5, test_data) 

# Add predicted values to the test data set
test_data$Predicted_Sale_Price <- predictions

# Show the difference in values
test_data$Predicted_vs_Actual <- round(test_data$Sale_Price - test_data$Predicted_Sale_Price,2)

# Evaluate metrics

metrics <- metrics(test_data, truth = Sale_Price, estimate = Predicted_Sale_Price)

metrics$.estimate <- round(metrics$.estimate,2)

test_data %>%
ggplot(aes(x = Sale_Price, y = Predicted_Sale_Price)) +
  geom_point(alpha = 0.6) + #Size of the points
  geom_abline(slope = 1, intercept = 0, color = "red") + #line to show slope
  labs(
    title = "Predicted vs Actual Sale Price",
    x = "Actual Sale Price",
    y = "Predicted Sale Price"
  ) +
  theme_minimal()
```

### Evaluate the Model
##lm_t1 = Neighborhood and Overall quality
##lm_t2 = Overall quality and Ground living area
##lm_t3 = Lot area and Garage car
```{r}

performance(lm_t1)
performance(lm_t4)
performance(lm_t5)

#tidy(lm_fit)

#parameters::model_parameters(lm_1)

performance::check_model(lm_t1)
performance::check_model(lm_t4)
performance::check_model(lm_t5)


```

Here is the checklist to evaluate a model from the "check_model()" method of the Performance library.

##Findings(Neighborhood and Overall quality):


-   **Posterior Predictive Check** - The prediction values align with the observed values for the most part but on the top there are some extreme values.


-   **Linearity** - Residuals are scatter randomly. Line is flat and horizontal with a bit of exception at higher value. There is no systematic pattern.


-   **Homogeneity of Variance (Homoscedacity)** - The spread of residuals is not consistent across fitted values. As is goes more extreme the residuals are increased slightly on higher fitted values.


-   **Influential Observations** - Most of the points are in the contour execpt one point at 463.


-   **Collinearity** - There is no strong collinearity. VIF is less than 5 which is best.


-   **Normality of Residuals** - The residuals are not normally distributed. Points are deviated from reference line.


##Findings(Overall quality and Ground living area):


-   **Posterior Predictive Check** - The prediction values align with the observed values for the most part but on the top there are some extreme values.


-   **Linearity** - Residuals are scatter randomly. Line is flat and horizontal for most part and is slightly downward at high value.


-   **Homogeneity of Variance (Homoscedacity)** - The spread of residuals are consistent across fitted values and consistently increases as in approaches  to higher value.


-   **Influential Observations** - Most of the points are in the contour expect some extreme values.


-   **Collinearity** -  VIF 1.5 for both indicating low collinearity and no significant collinearity among predictors.


-   **Normality of Residuals** - The residuals are not normally distributed. Points are deviated from reference line in the beginning and at the end.


##Findings(Lot area and Garage car):


-   **Posterior Predictive Check** - The prediction values doesn't align with the observed values at around 0-300k.


-   **Linearity** - Residuals are scatter randomly. Line doesn't align with the horizontal line.


-   **Homogeneity of Variance (Homoscedacity)** - The spread of residuals is not consistent across fitted values. As is goes more extreme the residuals line decreases and increases like funnel shape.


-   **Influential Observations** - There are multiple points that influences the contour and the most extreme is 798.


-   **Collinearity** - VIF is 1 for both indicating there are no strong collinearity among the predictors.


-   **Normality of Residuals** - The residuals are not normally distributed. Points are deviated from reference line at both tails.


### Check Variable Importance

Variable importance assesses which predictor variables that have the strongest predictive value to the outcome. In this case, we can make the following observations:

-   Gross Living Area was the number one predictor with neighborhood, Ground living area with overall quality, and Garage car with lot area.
```{r}
vip::vip(lm_all_clean, num_features = 20)
```

```{r}
vip::vip(lm_t1, num_features = 20)
```



```{r}

vip::vip(lm_t4, num_features = 20)

```
```{r}
vip::vip(lm_t5, num_features = 20)

```

